% ============================================================
% PI-SSL Paper
% Physics-Informed Self-Supervised Learning for
% Cross-Domain Wi-Fi Gesture Recognition
%
% STATUS (April 2026):
%   COMPLETE — backed by verified implementation and results:
%     Section 1  — Introduction
%     Section 2  — Related Work
%     Section 3  — Dataset and Preprocessing
%     Section 5.1 — Experimental Setup
%     Section 5.2 — Baseline Results (BVP+SVM, LeNet, CNN-GRU)
%
%   PLACEHOLDER — fill once PI-SSL is implemented and run:
%     Section 4  — Methodology (encoder, augmentations, objective)
%     Section 5.3 — PI-SSL and SimCLR results
%     Section 5.4 — Ablation Study
%     Section 6  — Conclusion (final paragraph)
%     Abstract   — replace XX.XX with final accuracy
%
% Template: IEEEtran (journal mode)
% Compile:  pdflatex pi_ssl_paper.tex
% ============================================================

\documentclass[journal]{IEEEtran}

% ── Packages ─────────────────────────────────────────────────────────────────
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{url}
\usepackage{cite}
\usepackage{array}
\usepackage{tabularx}
\usepackage{makecell}
\usepackage{enumitem}

% ── Convenience macros ────────────────────────────────────────────────────────
\newcommand{\bvp}{\mathbf{X}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}

% =============================================================================
\begin{document}
% =============================================================================

\title{Physics-Informed Self-Supervised Learning for\\
Cross-Domain Wi-Fi Gesture Recognition}

\author{Ghulam~Fatima,~\IEEEmembership{Student Member,~IEEE,}
        and~Zahid~Halim,~\IEEEmembership{Member,~IEEE}%
\thanks{G.~Fatima and Z.~Halim are with the Department of Information
Management, National Yunlin University of Science and Technology (YunTech),
Yunlin 640, Taiwan (e-mail: D11423005@yuntech.edu.tw).}}

\maketitle

% =============================================================================
% ABSTRACT
% =============================================================================
\begin{abstract}
Wi-Fi-based gesture recognition is a compelling contactless sensing
modality enabled by commodity 802.11 infrastructure, yet existing
supervised methods generalise poorly across physical environments
because Channel State Information (CSI) is highly sensitive to room
geometry and multipath propagation conditions.
The Body-coordinate Velocity Profile (BVP) transformation partially
decouples the gesture signal from environmental factors, achieving
89.7\% mean accuracy under a 3-fold Leave-One-Room-Out (LORO)
protocol with 100\% of available labels on the Widar3.0 dataset.
However, annotating Wi-Fi gesture data is laborious, motivating
label-efficient learning approaches.
We propose PI-SSL (\textbf{P}hysics-\textbf{I}nformed
\textbf{S}elf-\textbf{S}upervised \textbf{L}earning), a framework
that exploits the physics of BVP formation as a self-supervisory
signal to learn domain-invariant representations from unlabeled data.
Through a rigorous supervised baseline study under the same LORO
protocol with only 25\% of labels, we establish that the best
supervised baseline (Widar\_LeNet from SenseFi) achieves
$69.14 \pm 8.19\%$ mean accuracy --- a gap of more than 20 percentage
points below the 89.7\% supervised target.
PI-SSL addresses this gap by pre-training a shared encoder over the
full unlabeled BVP corpus before fine-tuning on the limited labeled
set, achieving \todo{XX.XX\%} mean accuracy with 25\% labels.
\end{abstract}

\begin{IEEEkeywords}
Wi-Fi sensing, gesture recognition, self-supervised learning,
contrastive learning, cross-domain generalisation,
Body-coordinate Velocity Profile, semi-supervised learning.
\end{IEEEkeywords}

% =============================================================================
\section{Introduction}
\label{sec:intro}
% =============================================================================

Wi-Fi-based gesture recognition leverages the 802.11 wireless
infrastructure already deployed in homes, offices, and public spaces
to detect and classify human hand and body movements without cameras,
wearables, or dedicated sensors.
Unlike vision-based systems, Wi-Fi sensing operates through walls,
is robust to lighting conditions, and does not record personally
identifiable imagery, making it a practically attractive modality for
smart-home control, healthcare monitoring, and human--computer
interaction~\cite{ma2023sensefi}.

The dominant technical signal is Channel State Information (CSI),
which encodes how the Wi-Fi channel transforms a transmitted signal
at the subcarrier level.
Human motion perturbs the multipath propagation of radio waves,
leaving characteristic imprints in the complex CSI coefficients.
However, CSI is highly sensitive to the physical environment: room
geometry, furniture arrangement, transmitter--receiver placement, and
reflection paths all shape the CSI baseline in ways unrelated to
the gesture.
As a result, classifiers trained in one room generalise poorly to a
new room, even when the same gesture is performed by the same person
under nominally similar conditions.
This \emph{domain shift} problem is the central obstacle to practical
cross-environment deployment of Wi-Fi gesture recognition systems.

The Body-coordinate Velocity Profile (BVP) was introduced by Zheng
\textit{et al.}~\cite{zheng2019widar3} as a principled representation
that partially decouples the gesture signal from environmental factors.
BVP is derived by projecting CSI dynamics into a coordinate system
fixed to the human body, normalising out the room-specific multipath
geometry.
A Support Vector Machine trained on BVP features achieves $89.7\%$
mean classification accuracy under a 3-fold Leave-One-Room-Out (LORO)
cross-validation on the Widar3.0 dataset~\cite{zheng2019widar3},
establishing the state-of-the-art supervised benchmark for
cross-room Wi-Fi gesture recognition.
This result, however, is obtained using the full labeled training corpus.

Label acquisition is a practical bottleneck for Wi-Fi gesture
recognition.
Each recording session requires a human participant to perform a
predefined gesture vocabulary in a controlled environment, and every
gesture instance must be individually verified and annotated.
The Widar3.0 dataset involved 17 participants across 14 session
environments, requiring substantial data collection effort.
In realistic deployment scenarios, only a fraction of recorded data may
be accompanied by verified labels, while the remainder is unannotated.
We therefore study the \emph{label-efficiency problem}: given only
25\% of the available labeled training data, can a model match or
exceed the $89.7\%$ supervised baseline that requires all labels?

To close this gap, we propose PI-SSL
(\textbf{P}hysics-\textbf{I}nformed \textbf{S}elf-\textbf{S}upervised
\textbf{L}earning), a two-stage framework that exploits the physics of
BVP formation as a self-supervisory signal.
The key insight is that the BVP transformation is derived from the
Fresnel zone model of Wi-Fi propagation, which encodes
domain-invariant constraints on how body velocity components project
onto the Doppler frequency spectrum.
These constraints can be operationalised as physically grounded
augmentation policies and pre-training objectives that steer the
learned encoder toward gesture-discriminative, environment-agnostic
representations --- without requiring any labels.
The pre-trained encoder is then fine-tuned using the small labeled
set, combining the representational capacity built over the full
unlabeled corpus with the supervisory signal from labeled examples.

The contributions of this paper are as follows:

\begin{itemize}[leftmargin=*]

  \item \textbf{Comprehensive baseline study.}
  We conduct a rigorous supervised baseline evaluation under the
  3-fold LORO protocol with 25\% labeled data on Widar3.0,
  evaluating BVP+SVM, Widar\_LeNet~\cite{ma2023sensefi}, and
  Widar\_CNN-GRU~\cite{ma2023sensefi}.
  We report per-fold and mean accuracy for both the 25\%-label and
  100\%-label settings, providing a transparent characterisation of
  the label-efficiency gap that PI-SSL targets.

  \item \textbf{Empirical analysis of cross-room evaluation.}
  We identify and quantify three systematic phenomena in the LORO
  benchmark: (i)~a 15--20 point accuracy drop on the hardest fold
  driven by a severe train/test size asymmetry; (ii)~a 6--10 point
  accuracy cost from reducing labels to 25\% across all architectures;
  and (iii)~a persistent $>$10-point gap between supervised baselines
  and the 89.7\% state-of-the-art even with 100\% labels, confirming
  that the domain gap cannot be closed by architecture alone.

  \item \textbf{PI-SSL framework.}
  We propose a physics-informed SSL pre-training scheme that uses
  BVP-derived domain-invariant constraints as self-supervisory signals,
  enabling the encoder to learn cross-room-generalisable representations
  from unlabeled data (Section~\ref{sec:method}).

  \item \textbf{Label-efficient cross-room recognition.}
  PI-SSL achieves \todo{XX.XX\%} mean accuracy under the standard-6
  LORO benchmark using only 25\% of labels, targeting the
  fully-supervised state-of-the-art of 89.7\%~\cite{zheng2019widar3}
  that requires 100\% labels.

\end{itemize}

% =============================================================================
\section{Related Work}
\label{sec:related}
% =============================================================================

\subsection{Wi-Fi Gesture Recognition}
\label{sec:related:wifi}

The feasibility of recognising human gestures from Wi-Fi CSI
measurements has been demonstrated across a broad range of gesture
vocabularies and sensing modalities over the past decade.
Early approaches extracted motion signatures from amplitude envelopes
or received signal strength (RSSI) variations, but subcarrier-level
CSI measurements provided by commodity 802.11n/ac network interface
cards proved substantially richer, capturing Doppler frequency shifts
and multipath changes at the scale of individual hand movements.

A decisive step toward cross-environment robustness was the
introduction of the BVP representation by Zheng
\textit{et al.}~\cite{zheng2019widar3}.
Rather than operating directly on the raw CSI, the BVP transformation
projects channel dynamics into a body-relative velocity space,
partially normalising out room-specific propagation geometry.
The resulting Widar3.0 system achieves 89.7\% mean accuracy under
the LORO protocol on a large-scale multi-room dataset, establishing
the primary benchmark against which we evaluate.

Ma \textit{et al.}~\cite{ma2023sensefi} introduced SenseFi, a
library and benchmark for deep-learning-based Wi-Fi human sensing.
SenseFi provides reference implementations of several architectures ---
including the Widar\_LeNet (CNN-5) and Widar\_CNN-GRU used in this
work --- alongside evaluations on multiple sensing datasets.
We adopt both architectures as supervised baselines, adapting them
from their original $T{=}22$ time-frame input to our $T^{*}{=}20$
preprocessing.
A key distinction is that SenseFi reports results on random
subject-independent splits across all 22 gesture classes, whereas
we evaluate exclusively under the 6-class LORO protocol,
providing a more stringent and directly comparable cross-domain
benchmark.

\subsection{Self-Supervised and Contrastive Learning}
\label{sec:related:ssl}

Self-supervised learning (SSL) defines auxiliary training objectives
whose supervisory signal is derived from the data itself, enabling
representation learning without human-annotated labels.
Contrastive methods learn by attracting embeddings of differently
augmented views of the same sample and repelling embeddings of
different samples.
SimCLR~\cite{chen2020simclr} demonstrated that a simple framework
combining strong data augmentations, a non-linear projection head,
and the normalised temperature-scaled cross-entropy (NT-Xent) loss
achieves state-of-the-art linear evaluation accuracy on large-scale
image benchmarks, establishing contrastive SSL as a practical
alternative to supervised pre-training.

The success of contrastive SSL depends critically on the augmentation
strategy: effective augmentations preserve semantically relevant
content while breaking superficial correlations specific to the
recording environment.
For natural images, random cropping, colour jitter, and Gaussian blur
achieve this; these operations, however, carry no semantic meaning
for BVP volumes.
Generic signal augmentations such as additive Gaussian noise, temporal
shift, and amplitude scaling introduce useful regularisation but do
not specifically encode the domain-invariant structure of BVP signals.
PI-SSL addresses this limitation by grounding augmentations in the
physics of BVP formation, selecting view-generation operations that
correspond to physical changes in transmitter--receiver geometry
while preserving the gestural velocity content.

\subsection{Semi-Supervised and Cross-Domain Wi-Fi Sensing}
\label{sec:related:semi}

Several recent works have addressed cross-environment transfer with
limited supervision in Wi-Fi sensing.
ARC-Fi~\cite{yang2023arcfi} applies semi-supervised contrastive
learning to achieve cross-room generalisation on Widar3.0,
reporting 79--91\% accuracy across varying configurations.
Direct numerical comparison with ARC-Fi against our LORO results is
approximate, as ARC-Fi uses a different gesture subset and split
strategy; we therefore cite it as a qualitative reference rather
than a head-to-head baseline.

In the broader signal processing literature, physics-informed
constraints have been incorporated into self-supervised and
semi-supervised objectives for radar target detection, sonar
classification, and wireless channel estimation, demonstrating that
physical domain knowledge can substantially improve label efficiency
in settings where data collection is costly.
To our knowledge, PI-SSL is the first framework to apply
BVP-physics-derived constraints specifically as SSL objectives for
cross-domain Wi-Fi gesture recognition.

% =============================================================================
\section{Dataset and Preprocessing}
\label{sec:dataset}
% =============================================================================

\subsection{Widar3.0 BVP Dataset}
\label{sec:dataset:overview}

We evaluate on the publicly released Widar3.0
dataset~\cite{zheng2019widar3}, which provides Body-coordinate
Velocity Profile (BVP) signals collected across multiple physical
environments using commodity Wi-Fi hardware.
Each BVP sample is derived from Channel State Information (CSI)
measurements by applying a domain transformation that expresses hand
motion as a velocity profile in a coordinate system fixed to the
human body, partially decoupling the signal from the specific
transmitter--receiver geometry of the recording environment.

The dataset comprises recordings across \textbf{14 session
environments} spanning three distinct physical rooms: a
\emph{Classroom} (Room~1, 5~sessions), a \emph{Hall} (Room~2,
8~sessions), and an \emph{Office} (Room~3, 1~session).
Seventeen participants (user1--user17) performed gesture sequences
across these sessions.
After filename parsing and gesture label resolution
(Section~\ref{sec:dataset:labels}), we identify 43{,}658 candidate
files.
Of these, 500 are discarded because they belong to users whose
gesture mappings are absent from the official dataset documentation
for three partially-recorded environments, yielding
\textbf{43{,}158 usable files} before preprocessing.

\subsubsection*{File format}
Each \texttt{.mat} file stores one gesture instance as a
three-dimensional array
\texttt{velocity\_spectrum\_ro} $\in \R^{20 \times 20 \times T}$,
where the first axis indexes Doppler-velocity bins ($20$ bins),
the second indexes spatial/angle bins ($20$ bins), and
$T \in [0,\,38]$ is the number of recorded time frames, which varies
across samples due to differences in gesture duration and recording
length.
The filename encodes all per-sample metadata according to the format:
\begin{equation*}
  \texttt{\{user\}-\{gesture\}-\{torso\}-\{face\}-\{rep\}-[\ldots]-L\{link\}.mat}
\end{equation*}
where \texttt{gesture} is a local integer label whose semantic meaning
is session-specific, as described next.

\subsection{Gesture Label Resolution}
\label{sec:dataset:labels}

A critical and underemphasised property of Widar3.0 is that the
gesture integers embedded in filenames are \emph{local to each
recording session}: the same integer maps to different semantic classes
across environments.
For example, local label~\texttt{1} encodes \textsc{Push\&Pull} in
the 20181109-VS session but \textsc{Draw-1} (a digit gesture) in
20181112-VS.
Treating raw filename integers as global class labels introduces
spurious cross-environment label noise, inflates the apparent number
of gesture classes, and would make standard-6 evaluation impossible.

We resolve every local label to a global semantic name using the
complete per-environment, per-user mapping published in the official
Widar3.0 documentation, yielding \textbf{21 unique semantic gesture
classes}.
The digit gesture \textsc{Draw-0} is absent from all recorded
sessions.
Table~\ref{tab:gestures} lists the full class inventory and the
per-class sample count after preprocessing.
Only four of the fourteen environments contain all six standard
benchmark gestures; these four environments are the sole contributors
to the supervised training and test sets used throughout this work.

\begin{table}[t]
  \centering
  \caption{Semantic gesture classes in Widar3.0 and sample counts after
           preprocessing. Standard benchmark gestures are marked
           ($\star$).}
  \label{tab:gestures}
  \renewcommand{\arraystretch}{1.15}
  \begin{tabular}{@{}llr@{}}
    \toprule
    \textbf{Group} & \textbf{Gesture} & \textbf{Samples} \\
    \midrule
    \multirow{6}{*}{Benchmark$^\star$}
      & Push\&Pull$^\star$    & 4{,}297 \\
      & Sweep$^\star$         & 4{,}174 \\
      & Clap$^\star$          & 4{,}170 \\
      & Slide$^\star$         & 4{,}875 \\
      & Draw-O-H$^\star$      & 3{,}748 \\
      & Draw-Zigzag-H$^\star$ & 3{,}750 \\
    \midrule
    \multirow{6}{*}{Extended}
      & Draw-N-H         & 3{,}124 \\
      & Draw-Triangle-H  & 2{,}999 \\
      & Draw-Rectangle-H & 3{,}000 \\
      & Draw-O-V         &    675  \\
      & Draw-Zigzag-V    & 1{,}925 \\
      & Draw-N-V         & 1{,}923 \\
    \midrule
    \multirow{9}{*}{Digit}
      & Draw-1 &   500 \\
      & Draw-2 &   500 \\
      & Draw-3 &   500 \\
      & Draw-4 &   500 \\
      & Draw-5 &   500 \\
      & Draw-6 &   492 \\
      & Draw-7 &   500 \\
      & Draw-8 &   500 \\
      & Draw-9 &   500 \\
    \midrule
    \multicolumn{2}{@{}l}{\textbf{Total}} & \textbf{43{,}152} \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Preprocessing}
\label{sec:dataset:preprocessing}

\subsubsection*{Per-sample normalisation}
BVP signal amplitude varies substantially across physical environments
due to differences in room geometry, path loss, and receiver
placement.
To remove environment-level amplitude offsets and encourage the
encoder to learn gesture kinematics rather than absolute signal
strength, we apply per-sample z-score normalisation to every BVP
volume before any other operation:
\begin{equation}
  \hat{\bvp} =
    \frac{\bvp - \mu(\bvp)}{\sigma(\bvp) + \varepsilon},
  \qquad \varepsilon = 10^{-6},
  \label{eq:zscore}
\end{equation}
where $\bvp \in \R^{20 \times 20 \times T}$ is the raw BVP array
and $\mu(\cdot)$, $\sigma(\cdot)$ denote the scalar mean and standard
deviation computed over all elements of the volume.
Normalisation is applied \emph{before} temporal padding so that the
appended zero frames carry the semantic of ``no signal'' and do not
influence the per-sample statistics.

\subsubsection*{Temporal alignment}
The time dimension $T$ varies from 0 to 38 frames across samples,
reflecting differences in gesture execution speed and recording
length.
We adopt $T^{*} = 20$ as the canonical sequence length, matching the
mode of the empirical $T$ distribution.
Samples with $T > T^{*}$ are truncated to the first $T^{*}$ frames,
retaining the kinematically rich onset and peak of the gesture
trajectory.
Samples with $T < T^{*}$ are zero-padded on the right to length
$T^{*}$.
Six files with $T = 0$ or corrupted array dimensionality are
excluded, yielding a final dataset of \textbf{43{,}152 samples},
each represented as a $20 \times 20 \times 20$ floating-point volume.

\subsubsection*{Storage and reproducibility}
The preprocessed dataset is stored as a single compressed NumPy
archive (\texttt{preprocessed.npz}, 159.6\,MB) containing BVP
volumes and all per-sample metadata (gesture class, room,
environment identifier, participant ID, repetition index).
Storing the entire corpus in a single memory-mapped archive
eliminates per-sample \texttt{.mat} parsing overhead during GPU
training --- reducing data access from approximately 1\,ms per sample
to a direct array index operation --- and ensures that every
train/test split is deterministically reproducible from a fixed
random seed.

% =============================================================================
\section{Methodology}
\label{sec:method}
% =============================================================================

\subsection{Framework Overview}
\label{sec:method:overview}

PI-SSL follows a two-stage training procedure.
In the first stage, a shared convolutional encoder $f_\theta$ is
pre-trained on the full unlabeled BVP corpus from the training rooms
using a physics-informed contrastive objective; no gesture labels are
used.
In the second stage, the pre-trained encoder is augmented with a
linear classification head and fine-tuned using only the 25\%
labeled subset of standard-6 training files.

The motivation for this design is empirical: our supervised baseline
analysis (Section~\ref{sec:experiments:results}) shows that neural
networks trained directly on the small labeled set overfit severely ---
Widar\_LeNet reaches 100\% training accuracy by epoch~20 while
achieving only 57--75\% test accuracy.
Pre-training on the full unlabeled pool ($\sim$11k--40k samples per
fold, Table~\ref{tab:splits}) provides the encoder with a
substantially richer basis for generalisation before the supervised
signal is introduced.

\subsection{Encoder Architecture}
\label{sec:method:encoder}

\todo{Describe the shared encoder backbone. Planned: 3-D convolutional
network operating on the $(1 \times 20 \times 20 \times 20)$
channel-first BVP volume, producing a $d$-dimensional embedding via
global average pooling. Specific layer counts, channel widths, and
normalisation strategy to be determined.}

\subsection{Physics-Informed Augmentations}
\label{sec:method:augmentations}

\todo{Describe the physically motivated view-generation policy.
Candidate augmentations include: (i)~Doppler-axis velocity scaling,
corresponding to changes in gesture execution speed;
(ii)~spatial-axis perturbation, corresponding to changes in the
transmitter--receiver orientation; (iii)~temporal cropping within the
gesture window; (iv)~calibrated additive noise. Physical justification
for each operation in terms of the BVP Fresnel zone model to be
detailed here.}

\subsection{Pre-Training Objective}
\label{sec:method:objective}

\todo{The NT-Xent (normalised temperature-scaled cross-entropy) loss
from SimCLR~\cite{chen2020simclr} is used as the contrastive
objective, applied over a two-layer MLP projection head $g_\phi$
that maps the encoder embedding to a 128-dimensional normalised
vector. Temperature $\tau$, batch size, and number of pre-training
epochs to be specified once hyperparameter search is complete.}

\subsection{Supervised Fine-Tuning}
\label{sec:method:finetuning}

\todo{Describe the fine-tuning protocol: replacement of the projection
head with a linear classifier, partial or full unfreezing of
$f_\theta$, regularisation strategy (dropout, weight decay), learning
rate schedule, and early stopping criterion.}

% =============================================================================
\section{Experiments}
\label{sec:experiments}
% =============================================================================

\subsection{Experimental Setup}
\label{sec:experiments:setup}

\subsubsection*{Evaluation protocol}
We follow the Leave-One-Room-Out (LORO) cross-validation protocol
from Zheng~\textit{et al.}~\cite{zheng2019widar3}, the established
benchmark for cross-environment Wi-Fi gesture recognition.
In each of three folds, one physical room is withheld as the test
domain and the remaining two rooms form the training domain.
No environment seen during training appears during evaluation,
providing a stringent measure of cross-domain generalisation.
Evaluation is restricted to the six standard benchmark gestures
(\textsc{Push\&Pull}, \textsc{Sweep}, \textsc{Clap}, \textsc{Slide},
\textsc{Draw-O-H}, \textsc{Draw-Zigzag-H}) to permit direct
numerical comparison with all published results on this dataset.

\subsubsection*{Semi-supervised split}
We simulate a low-annotation regime by designating \textbf{25\%} of
the standard-6 training files as the labeled set for supervised
fine-tuning.
The split is stratified by gesture class so that every class
contributes a proportional number of labeled examples, with a minimum
of one example per class.
The remaining 75\% of standard-6 training files are treated as
unlabeled and are available for the unlabeled pool in SSL pre-training
or consistency-based auxiliary losses.
SSL pre-training additionally uses all files from the training rooms
across all 21 gesture classes with labels ignored, maximising the
volume of Wi-Fi BVP signal exposed to the encoder.

Table~\ref{tab:splits} reports the exact split sizes for each fold.

\begin{table}[t]
  \centering
  \caption{Dataset split sizes per LORO fold (verified against actual
           training runs). The SSL pre-train set spans all 21 gesture
           classes; labeled and unlabeled sets contain standard-6
           gestures only.}
  \label{tab:splits}
  \renewcommand{\arraystretch}{1.15}
  \begin{tabular}{@{}clrrrr@{}}
    \toprule
    \textbf{Fold} & \textbf{Test Room} &
    \textbf{SSL Pre-train} & \textbf{Labeled} &
    \textbf{Unlabeled} & \textbf{Test} \\
    \midrule
    0 & Office    & 40{,}157 & 5{,}505 & 16{,}514 &  2{,}995 \\
    1 & Classroom & 11{,}744 & 2{,}155 &  6{,}463 & 16{,}396 \\
    2 & Hall      & 34{,}410 & 4{,}848 & 14{,}543 &  5{,}623 \\
    \bottomrule
  \end{tabular}
\end{table}

A structurally important feature of the Widar3.0 LORO benchmark is
the pronounced asymmetry across folds.
Fold~1 (Classroom) has the fewest training environments but the
largest test set (16{,}396 samples), while Fold~0 (Office) has a
small test set (2{,}995 samples) drawn from a single recording
session.
This asymmetry means that Fold~1 is consistently the hardest fold
across all methods, and reported standard deviations are dominated
by Fold~1 performance.
We report per-fold accuracy for all methods to make this asymmetry
transparent.

\subsubsection*{Training details for supervised baselines}
All supervised neural-network baselines are trained for 100 epochs
using the Adam optimiser with initial learning rate $10^{-3}$ and
weight decay $10^{-4}$.
Mini-batch size is 64 for all models.
No data augmentation is applied during supervised training to provide
a clean lower bound on supervised performance and an unambiguous
comparison with PI-SSL, which benefits from augmentation during
pre-training.
Random seed 42 is fixed for all data splits and weight initialisations
to ensure full reproducibility.

\subsubsection*{Evaluation metric}
Performance is reported as per-fold top-1 classification accuracy
and mean accuracy $\pm$ standard deviation across the three folds.
The target for PI-SSL is ${>}89.5\%$ mean accuracy --- exceeding the
Widar3.0 supervised reference of $89.7\%$~\cite{zheng2019widar3}
while using only 25\% of the available labels.

\subsubsection*{Baselines}
Three baseline systems are evaluated under the 25\%-label LORO
protocol:

\begin{itemize}[leftmargin=*]

  \item \textbf{BVP+SVM.} An RBF-kernel Support Vector Machine
  ($C = 10$, $\gamma = \texttt{scale}$) trained on the raw flattened
  BVP volume ($20 \times 20 \times 20 \to 8{,}000$-dimensional
  vector) with per-fold \texttt{StandardScaler} normalisation.
  This is the architecturally simplest baseline and establishes a
  lower bound on the performance of learned representations.
  Note that the Widar3.0 reference result of
  $89.7\%$~\cite{zheng2019widar3} uses hand-crafted BVP statistics
  rather than raw flattened volumes and a different train/test split;
  our BVP+SVM is an independent implementation and is not a
  reproduction of that result.

  \item \textbf{Widar\_LeNet}~\cite{ma2023sensefi}. The five-layer
  CNN (CNN-5) from the SenseFi benchmark, originally designed for
  $T{=}22$ input frames.
  We adapt the number of input channels from 22 to 20 to match our
  $T^{*}{=}20$ preprocessing.
  The model treats the $T$ time frames as channels to a 2-D
  convolutional stack ($32 \to 64 \to 96$ feature maps, kernel sizes
  $6 \to 3 \to 3$), learning spatial--velocity features jointly
  across all time steps, followed by two fully connected layers.

  \item \textbf{Widar\_CNN-GRU}~\cite{ma2023sensefi}. A per-frame
  CNN encoder combined with a single-layer GRU temporal model from
  SenseFi, adapted to $T{=}20$.
  The final \texttt{Softmax} layer is removed to be compatible with
  PyTorch's \texttt{CrossEntropyLoss}, which expects raw logits.
  Each of the $T$ frames is independently encoded by a shared 2-D
  CNN ($1 \to 8 \to 16$ channels, with $2\times$ max-pooling), the
  resulting 64-dimensional frame embeddings are fed sequentially
  into the GRU, and the final hidden state is passed to a linear
  classifier.

\end{itemize}

\subsection{Baseline Results}
\label{sec:experiments:results}

Table~\ref{tab:results} reports per-fold and mean classification
accuracy for all methods.
Rows marked with $\dagger$ show the same architecture trained on
100\% labels, providing an internal upper bound and directly
quantifying the label-efficiency gap for each model.

\begin{table*}[t]
  \centering
  \caption{Classification accuracy (\%) on Widar3.0 standard-6
           gestures, 3-fold LORO. All PI-SSL and baseline entries use
           25\% of training labels unless marked $\dagger$ (100\%
           labels). $^*$~Widar3.0 TPAMI uses hand-crafted BVP
           statistics and a different evaluation setup; not directly
           comparable to our baselines (see text). Best 25\%-label
           result among our baselines is in \textbf{bold}.}
  \label{tab:results}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    \textbf{Method} &
    \makecell{\textbf{Fold 0}\\\textit{Office}} &
    \makecell{\textbf{Fold 1}\\\textit{Classroom}} &
    \makecell{\textbf{Fold 2}\\\textit{Hall}} &
    \textbf{Mean $\pm$ Std} \\
    \midrule
    \multicolumn{5}{@{}l}{\textit{Supervised upper bounds (100\% labels, for reference only)}} \\[2pt]
    BVP+SVM$^\dagger$                             & 77.36 & 58.69 & 71.05 & 69.03 $\pm$ 7.76 \\
    Widar\_LeNet$^\dagger$~\cite{ma2023sensefi}   & 85.58 & 68.74 & 83.14 & 79.15 $\pm$ 7.43 \\
    Widar\_CNN-GRU$^\dagger$~\cite{ma2023sensefi} & 79.87 & 63.84 & 78.20 & 73.97 $\pm$ 7.19 \\
    Widar3.0 TPAMI~\cite{zheng2019widar3}$^*$     & \multicolumn{3}{c}{---} & 89.70 \\[4pt]
    \midrule
    \multicolumn{5}{@{}l}{\textit{25\% labeled training data}} \\[2pt]
    BVP+SVM (ours)                                & 73.29 & 51.62 & 62.87 & 62.59 $\pm$ 8.85 \\
    Widar\_LeNet~\cite{ma2023sensefi}             & 75.46 & 57.58 & 74.39 & \textbf{69.14 $\pm$ 8.19} \\
    Widar\_CNN-GRU~\cite{ma2023sensefi}           & 74.72 & 53.67 & 70.46 & 66.28 $\pm$ 9.09 \\
    SimCLR + fine-tune                            & --    & --    & --    & -- \\
    \midrule
    \textbf{PI-SSL (ours)}                        & --    & --    & --    & \todo{\textbf{XX.XX}} \\
    \bottomrule
  \end{tabular}
\end{table*}

\subsubsection*{BVP+SVM}
The BVP+SVM baseline achieves $62.59 \pm 8.85\%$ with 25\% labels
and $69.03 \pm 7.76\%$ with 100\% labels.
The 21.7-point spread across folds reflects the structural asymmetry
of the LORO benchmark: Fold~1 (Classroom) provides only 2{,}155
labeled training samples for 16{,}396 test samples, while Fold~0
(Office) has a more favourable ratio of 5{,}505 training to 2{,}995
test samples.
The Classroom fold consistently produces the lowest accuracy across
all methods, making it the dominant driver of inter-fold variance.

Per-gesture accuracy averaged over all three folds (25\% labels)
shows that \textsc{Clap} is the most distinguishable class
($77.94\%$), consistent with its impulsive signal pattern, while
\textsc{Draw-O-H} ($44.29\%$) and \textsc{Slide} ($46.85\%$) are
the most challenging, reflecting the fine-grained spatial
trajectories and subtle Doppler profiles of those motions.

\subsubsection*{Widar\_LeNet}
Widar\_LeNet achieves $69.14 \pm 8.19\%$ with 25\% labels and
$79.15 \pm 7.43\%$ with 100\% labels --- improvements of $+6.6$
and $+10.1$ percentage points over BVP+SVM, confirming that
convolutional feature learning over the spatial--velocity plane is
substantially more informative than a flattened BVP vector for
cross-room generalisation.

A notable finding is that training accuracy reaches $100\%$ before
epoch~20 in every fold, yet test accuracy plateaus between $57.6\%$
and $75.5\%$, indicating severe overfitting under the 25\%-label
regime.
The train--test accuracy gap of $25$--$42$ percentage points
quantifies the mismatch between the small labeled training
distribution and the held-out test room.
This motivates the SSL pre-training strategy of PI-SSL: exposing the
encoder to the full unlabeled BVP corpus before fine-tuning provides
a substantially richer basis for cross-domain generalisation.

\subsubsection*{Widar\_CNN-GRU}
Widar\_CNN-GRU achieves $66.28 \pm 9.09\%$ with 25\% labels and
$73.97 \pm 7.19\%$ with 100\% labels.
Despite adding explicit temporal modelling through the GRU layer,
CNN-GRU underperforms Widar\_LeNet by $2.9$ percentage points at
25\% labels.
Examination of the training curves reveals the mechanism: CNN-GRU
converges to $88$--$91\%$ training accuracy without saturating,
while Widar\_LeNet saturates at $100\%$ before epoch~20.
The recurrent parameters of the GRU require more labeled examples
to train effectively, making the architecture less data-efficient
under the 25\%-label constraint.
At 100\% labels the gap narrows to $5.2$ points, consistent with
CNN-GRU's greater data hunger.
Notably, CNN-GRU trains $11\times$ faster than Widar\_LeNet
(12.7~min vs.\ 144.8~min for all three folds), owing to the lighter
per-frame CNN encoder.

\subsubsection*{Summary and implications for PI-SSL}
Three consistent patterns emerge from the supervised baseline
results:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Fold asymmetry.} Every method suffers a $15$--$20$
  point accuracy drop on Fold~1 (Classroom) relative to Folds~0
  and~2, driven by the LORO protocol's train/test size imbalance.
  This is a property of the benchmark, not of any model; we surface
  it explicitly because it dominates reported standard deviations.
  \item \textbf{Label-efficiency gap.} Reducing labels from 100\% to
  25\% costs $6$--$10$ accuracy points for every architecture, with
  the gap widening for architecturally more complex models.
  \item \textbf{Supervised ceiling.} The best supervised baseline with
  100\% labels ($79.15\%$) falls $10.6$ points below the Widar3.0
  TPAMI reference of $89.7\%$, confirming that the cross-room domain
  gap cannot be closed by architecture choice alone.
\end{enumerate}

These findings directly motivate PI-SSL: the domain gap requires
learning representations that are invariant to room-specific factors,
and the label-efficiency gap requires leveraging the large unlabeled
corpus --- both of which are addressed by physics-informed SSL
pre-training.

\subsection{PI-SSL Results}
\label{sec:experiments:pissl}

\todo{Report per-fold accuracy, mean, and standard deviation for
PI-SSL and SimCLR+fine-tune once training is complete. Fill the
corresponding rows in Table~\ref{tab:results}.}

\subsection{Ablation Study}
\label{sec:experiments:ablation}

\todo{Report ablation results studying the contribution of each
component: (i)~SSL pre-training with generic augmentations only,
no physics grounding; (ii)~physics-informed augmentations applied
during supervised training only, no SSL pre-training; (iii)~full
PI-SSL combining physics-informed SSL pre-training with supervised
fine-tuning.}

% =============================================================================
\section{Conclusion}
\label{sec:conclusion}
% =============================================================================

We have presented PI-SSL, a physics-informed self-supervised learning
framework for cross-domain Wi-Fi gesture recognition.
A comprehensive supervised baseline evaluation on the Widar3.0 dataset
under the 3-fold LORO protocol with 25\% labeled data establishes
three empirical findings: the best supervised baseline achieves
$69.14\%$ mean accuracy, leaving a gap of more than 20 points below
the state-of-the-art obtained with 100\% labels; severe overfitting
under the 25\%-label regime is the dominant failure mode for all
neural-network baselines; and the Classroom fold's structural
train/test asymmetry drives inter-fold variance across all methods.
PI-SSL addresses these challenges by pre-training a shared encoder
over the full unlabeled BVP corpus using physics-informed contrastive
objectives before supervised fine-tuning on the limited labeled set.

\todo{Complete once PI-SSL results are available: report final
accuracy, compare to baselines and state-of-the-art, discuss which
gesture classes benefit most, and state limitations and future work.}

% =============================================================================
% References
% =============================================================================

\begin{thebibliography}{9}

\bibitem{zheng2019widar3}
Y.~Zheng, Y.~Zhang, K.~Qian, G.~Zhang, Y.~Liu, C.~Wu, and Z.~Yang,
``Zero-Effort Cross-Domain Gesture Recognition with Wi-Fi,''
in \emph{Proc. 17th ACM Int. Conf. Mobile Systems, Applications, and
Services (MobiSys)}, Seoul, South Korea, Jun.\ 2019, pp.~313--325.

\bibitem{ma2023sensefi}
Y.~Ma, G.~Zhou, and S.~Wang,
``SenseFi: A Library and Benchmark on Deep-Learning-Empowered Wi-Fi
Human Sensing,''
\emph{Patterns}, vol.~4, no.~3, p.~100703, Mar.\ 2023.

\bibitem{yang2023arcfi}
H.~Yang \textit{et al.},
``ARC-Fi: Adaptive and Robust Cross-Environment Wi-Fi Sensing via
Semi-Supervised Contrastive Learning,''
\emph{arXiv preprint arXiv:2310.06328}, Oct.\ 2023.

\bibitem{chen2020simclr}
T.~Chen, S.~Kornblith, M.~Norouzi, and G.~Hinton,
``A Simple Framework for Contrastive Self-Supervised Learning,''
in \emph{Proc. 37th Int. Conf. Machine Learning (ICML)},
Virtual, Jul.\ 2020, pp.~1597--1607.

\end{thebibliography}

\end{document}
